---
title: "File Preparation for GSEA"
author: "Paulo Magalang"
date: "3/31/2022"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
# load in libraries
library(tidyverse)
library(stringr)
library(dplyr)
```

```{r}
# load in data
blastp_dnds <- read_tsv(file = "../output/TanOak_blastp_orthologs_dnds.tsv")

head(blastp_dnds)
```

```{r}
# drop rows with non-finite dn/ds ratios and missing GO annotations
blastp_dnds_filtered <- blastp_dnds %>% filter(!is.nan(dnds), !is.na(dnds), !is.infinite(dnds), dnds >= 0, !is.na(GO_annotations))

# plot distribution of dn/ds ratios
blastp_dnds_filtered %>% ggplot(aes(x = dnds)) + geom_histogram(binwidth = 0.05) + 
  geom_vline(xintercept = 0.0876, color = "red") + geom_vline(xintercept = 1.11, color = "red")

blastp_dnds_filtered$dnds %>% summary()
```


File prep for GSEA-Preranked:

* gene sets (get all gene names associated with a GO term) as .gmt
  - 1st column: GO term
  - 2nd column: NAs
  - the rest: genes associated with GO term
  
* ranked list
  - GO term
  - dn/ds ratio


```{r}
# data wrangling for gene sets

# take entire column of GO terms and vectorize, filter out duplicate terms
all_go_terms <- str_extract_all(blastp_dnds_filtered$GO_annotations, "GO:[0-9]*") %>% unlist() %>% unique()

# init. gene sets, start with all gene names in one column for now
gene_sets <- tibble(all_go_terms) %>% mutate(desc = NA, genes = NA)

# for each GO term, check blastp_dnds_filtered for the gene(s) associated with the GO term
for (i in 1:dim(gene_sets)[1]) {
  # use logical vector to subset tanoak gene names, store in current index
  occurrences <- blastp_dnds_filtered$GO_annotations %>% str_detect(all_go_terms[i]) # indices should match up with gene_sets
  gene_sets[i, 3] <- blastp_dnds_filtered$tanoak[occurrences] %>% paste0(collapse = ",")
}

# can write out as .csv file and use sed or tr to change "," to "\t", save as .gmt file
#write.table(gene_sets, file = "../input/gsea/blastp/TanOak_blastp_geneset.csv", quote = FALSE, sep = ",", row.names = FALSE, col.names = FALSE)
```

```{bash, eval = FALSE}
cat TanOak_blastp_geneset.csv | sed 's/,/\t/g' > TanOak_blastp_geneset.gmt
```

```{r}
# ranked list
rank_list <- blastp_dnds_filtered %>% dplyr::select(tanoak, dnds)

names(rank_list) <- c("NAME", "dnds")

head(rank_list)

#write.table(rank_list, file = "../input/gsea/blastp/TanOak_blastp_preranked.rnk", quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)
```



File prep for GSEA (not used since a .chip file cannot be generated)

* expression dataset (use dn/ds ratio here) as .txt
  - 1st col: gene names (NAME)
  - 2nd col: NAs (DESCRIPTION)
  - 3rd col: dnds ratios

* phenotype lablels (0/1 in extreme_dnds column) as .cls

```{r}
# "expression" dataset
#expr <- blastp_dnds_filtered %>% select(tanoak, dnds) %>% mutate(DESCRIPTION = NA) %>%
#  select(tanoak, DESCRIPTION, dnds)
#
#names(expr) <- c("NAME", "DESCRIPTION", "dnds")
#
#head(expr)

#write.table(expr, file = "../input/gsea/blastp/TanOak_blastp_expression.txt", quote = FALSE, sep = "\t", row.names = FALSE)
```


```{r}
######
# can take advantage of indices since the values are already sorted, find associated dn/ds values

# 5736 * 0.05 = 286.8 -> top/bottom 287 genes

# establishing dn/ds cutoffs using row indices
# 1 for top/bottom 5% genes arranged by descending dn/ds values (<= 0.0876 and >= 1.11), 0 for everything else
# blastp_dnds_filtered %>% mutate(class = ifelse(dnds <= 0.0876 | dnds >= 1.11, 1, 0)) 

#ifelse(blastp_dnds_filtered$dnds <= 0.0876 | blastp_dnds_filtered$dnds >= 1.11, 1, 0) %>% sum() 
# 580 "extreme" genes
```

```{r}
# phenotype labels
#labels <- ifelse(blastp_dnds_filtered$dnds <= 0.0876 | blastp_dnds_filtered$dnds >= 1.11, 1, 0)
#write.table(matrix(labels, nrow = 1), file = "../input/gsea/blastp/TanOak_blastp_phenotype.cls", sep = "\t", row.names = FALSE, col.names = FALSE)

# write in first two lines using vim
# 5039 2 1
# # REG EXTR (0 = REG, 1 = EXTR)
# the labels
```



