---
title: "Explore Polysaccharide Binding"
output: html_notebook
---

Many of the TanOak genes coming up with high dN/dS in the polysaccharide binding category are LRK10L2 homologs.  The goal here is to align the TanOak paralogs with other related homologs, build a tree, and see where changes are happening.

Previously I blasted against the NR database, and that was a bit of a mess for building a tree.  Will try again, but with BLAST against selected organisms.

```{r}
library(tidyverse)
library(Biostrings)
library(UpSetR)
library(gplots)
```


## Sequence retrieval

Start with blastp of `../output/polysaccharide_peps.fa`

Used phyotozome blastp tool against: 

* Athaliana_TAIR10_167.proteome; 
* Qrubra_v2_1_687.proteome;
* Cdentata_v1_1_673.proteome; 
* Gmax_Wm82_a4_v1_508.proteome;
* Ptrichocarpa_v4_1_533.proteome

Job is at https://phytozome-next.jgi.doe.gov/blast-results/589527

Results table, retrieved peptides and cDNAs are in `output/blastp/polysaccharide_Seq_downloads/phytozome_blastp


## Sequence overlap

Have common sequences been retrieved for the different queries?

```{r}
hittable <- read_tsv("../output/polysaccharide_seq_downloads/phytozome_blastp/phytozome_polysach_blastp_071522.tsv",
                     name_repair = make.names) %>%
  select(-1) %>%
  mutate(fasta.name = str_c({str_replace(Species, " ", "_") %>% str_replace("\\.", "") },
                            str_extract(Protein, "[A-Za-z]*\\.?[A-Za-z0-9]*"),
                            sep="|") ) 

head(hittable)
```

```{r, fig.width=8}
hit.counts <- hittable %>% select(Protein, Query.ID) %>%
  unique() %>%
  table() %>%
  as.data.frame.matrix() 

upset(hit.counts, nsets = 10, nintersects = 50)
```

70100 (Wall-associated receptor kinase galacturonan-binding; Wall-associated receptor kinase C-terminal; LRK10L3),  
445 (Wall-associated receptor kinase galacturonan-binding; Wall-associated receptor kinase C-terminal); NA
Are unique:

585 (NA; LRK10L2) and   
804 (Wall-associated receptor kinase galacturonan-binding; ATL21) can be grouped together

781 (EGF-like domain profile.; Wall-associated receptor kinase galacturonan-binding; Serine/Threonine protein kinases active-site signature.; Protein kinase domain; Wall-associated kinase; EGF_CA; Protein kinase domain profile. ;WAK5	)  AND  
74803 (Calcium-binding EGF-like domain signature.; EGF-like domain profile.; Wall-associated receptor kinase galacturonan-binding; Calcium-binding EGF domain; EGF_CA; EGF-like domain signature 2.; Aspartic acid and asparagine hydroxylation site.; WAK2	)  can be grouped together

70081 (NA; ATL21),   
72558 (Wall-associated receptor kinase galacturonan-binding; ATL21), and   
833 (Wall-associated receptor kinase galacturonan-binding; ATL21), AND
74056 (NA; ATL96) can be grouped together

## Qrob homologs

Do blast to get all Qrob homologs

```{bash, eval=FALSE}
cd ../input/blast
blastp -query ../../output/polysaccharide_peps.fa -db Qrob_PM1N_CDS_aa_20161004.fa -outfmt 7 > ../../output/polysaccharide_Qrob_blast.tsv
```

```{r}
header <- str_split("query acc.ver, subject acc.ver, % identity, alignment length, mismatches, gap opens, q. start, q. end, s. start, s. end, evalue, bit score", pattern = ", ") %>%
  unlist() %>%
  make.names() %>%
  str_replace(fixed(".."), ".") %>%
  str_replace("X", "pct")

Qrob_hittable <- read_tsv("../output/polysaccharide_Qrob_blast.tsv", comment = "#", col_names = header)

Qrob_hittable
```


## 804 and 585

Let's start with 804 and 585.  Need to get TanOak proteins, Oak proteins, and the blast results in one file.  Then align and make tree.

```{r}
#table of 804 and 585 hits 
hits_804_585 <- hittable %>%
  filter(Query.ID %in% c("augustus_masked-585_pilon-processed-gene-3.11-mRNA-1",
                         "augustus_masked-804_pilon-processed-gene-0.1-mRNA-1")) %>%
  filter(!duplicated(fasta.name))

#blast results
polysach_peps <- readAAStringSet("../output/polysaccharide_seq_downloads/phytozome_blastp/phytozome_polysach_blastp_pep_071522.fa.gz")

homologs_804_585 <- polysach_peps[hits_804_585$fasta.name]

#add tanoak
tanoak_poly_aa <- readAAStringSet("../output/polysaccharide_peps.fa")

homologs_804_585 <- c(tanoak_poly_aa[c("augustus_masked-585_pilon-processed-gene-3.11-mRNA-1",
                                       "augustus_masked-804_pilon-processed-gene-0.1-mRNA-1")],
                      homologs_804_585)


#to get the oak protein orthologs, load in the blastp table

blastp_ortho <- read_csv("../output/tanoak_oak_blastp_orthologs.csv")

# filter for those that we want
oak_804_585_orthos <- blastp_ortho %>%
  filter(str_detect(query_id, "augustus_masked-585_pilon-processed-gene-3\\.11-mRNA-1|augustus_masked-804_pilon-processed-gene-0\\.1-mRNA-1"))

#read the Oak AA sequences and fix names
qrobAA <- readAAStringSet("../input/blast/Qrob_PM1N_CDS_aa_20161004.fa")
names(qrobAA) <- str_remove(names(qrobAA), " .*") #get rid of extra space and numbers

qrobAA_804_585_orthtos <- qrobAA[oak_804_585_orthos$subject_id]

#add "ortho" to name to make it clear
names(qrobAA_804_585_orthtos) <- str_c(names(qrobAA_804_585_orthtos), "_orttho")

#add oak
homologs_804_585 <- c(qrobAA_804_585_orthtos,
                      homologs_804_585)

homologs_804_585
```

now add other Qrob orthologs:
```{r}
# get them:
Qrob_hits_804_805 <- Qrob_hittable %>%
  filter(query.acc.ver %in% c("augustus_masked-585_pilon-processed-gene-3.11-mRNA-1",
                              "augustus_masked-804_pilon-processed-gene-0.1-mRNA-1"),
         !duplicated(subject.acc.ver),
         bit.score >= 100) %>%
  pull(subject.acc.ver)

qrob_804_585_homologs <- qrobAA[Qrob_hits_804_805]

homologs_804_585 <- c(homologs_804_585, qrob_804_585_homologs)
```


write out the table
```{r}
names(homologs_804_585) <- str_remove(names(homologs_804_585), ".*\\|")
writeXStringSet(homologs_804_585, "../output/tanoak_804_585_orthos.fa")
```

align!
```{bash, eval=TRUE}
time mafft --thread 3 --localpair --maxiterate 1000 --reorder ../output/tanoak_804_585_orthos.fa > ../output/tanoak_804_585_orthos_aligned.fa
```

build tree
```{bash, eval=TRUE}
cd ../output
fasttree tanoak_804_585_orthos_aligned.fa > tanoak_804_585_orthos.tre
```

## Export Tanoak, oak, and chestnut sequences:

```{r}
#585
homologs_804_585[c("augustus_masked-585_pilon-processed-gene-3.11-mRNA-1", "Qrob_P0680550.2", "Caden.12G048400", )] %>%
  writeXStringSet("../output_585_orthos_small.fa")

#804
homologs_804_585[c("augustus_masked-804_pilon-processed-gene-0.1-mRNA-1","Qrob_P0468790.2", "Caden.12G048800")] %>%
  writeXStringSet("../output_804_orthos_small.fa")
```

```{bash}
time mafft --thread 3 --localpair --maxiterate 1000 ../output_585_orthos_small.fa > ../output_585_orthos_small_aligned.fa
```

```{bash}
time mafft --thread 3 --localpair --maxiterate 1000 ../output_804_orthos_small.fa > ../output_804_orthos_small_aligned.fa
```

